Making Sequential Programs Deterministic

### 2019-04-04 design.tex
Edits in section 2.4.2

  $\forallt{\beta} \<Set>\angles{\<Det>\ \beta} \rightarrow \<NonDet>\ \beta$

  Basetype polymorphism is rarely used on its own.  Full type polymorphism is more
  common, even if the type is decomposed and the parts used independently,
  as in this more general signature for the \<choose> function:

  $\forallt{\kappa\ \beta} \<Set>\angles{\kappa\ \beta} \rightarrow \<NonDet>\ \beta$

\item
Qualifier polymorphism is common.  For example, here is the signature for
the \<length> method on strings:

@@

\noindent
and here is that for the addition operation:

@@

$\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$

$\<Det>\ \<String> \rightarrow \<Det> \<Int>$

The body of \<length> must typecheck at every possible instantiation of its
polymorphic function type.  At a use site of \<length>,
the most specific applicable instantiation is chosen.
Here is an example:
\todo{(put it in a table)}

\begin{Verbatim}
NonDet Int ni;
   Det Int di;

plus(ni, ni)  // has type NonDet Int; most specific applicable instantiation is NonDet Int -> NonDet Int
plus(di, di)  // has type Det Int; most specific applicable instantiation is Det Int -> Det Int
plus(ni, di)  // has type NonDet Int; most specific applicable instantiation is NonDet Int -> NonDet Int
\end{Verbatim}

We adopt the convention that  polymorphism is not instantiated in ways
that would create invalid types.  For example, the \<plus> polymorphic
function would not be instantiated at $\kappa = \<OrderNonDet>.$

@@

\noindent
because such an instantiation would include the invalid return type \<OrderNonDet Int>.
This means that the only two instantiations are at 
$\kappa = \<NonDet>$ and $\kappa = \<Det>$.  An invocation on a
\<OrderNonDet> collection would choose the \<NonDet> instantiation and the
size would be \<NonDet Int>.
%
A similar problem occurs for the \<first> operation that returns the first
element of a list.

\OurTypeSystem resolves this problem by introducing two operators over
polymorhic (type) variables, $\uparrow$ and $\downarrow$.
The $\uparrow$ operator converts \<OrderNonDet> to \<NonDet> and leaves the
other qualifiers unchanged.  The upward-pointing arrow is mnemonic for replacing
\<OrderNonDet> by something higher in the type hierarchy.
The $\downarrow$ operator is analogous, but converts \<OrderNonDet> to
\<Det>, which is lower in the type hierarchy.
\Cref{fig:poly-resolutions} formalizes their behavior.

@@

\todo{As I mentioned in email, $\<NonDet>\ \beta_e$ looks wrong to me,
  because $\beta_e$ may be a collection whose type arguments are not
  compatible with \<NonDet>.  Maybe just remark that the instantiation does
  not happen in that case.  If we try to adjust the type arguments to fit
  \<NonDet>, then unsoundness occurs.}

>>>>

  $\<choose> : \forallt{\beta} \<Set>\angles{\<Det>\ \beta} \rightarrow \<NonDet>\ \beta$

  Basetype polymorphism is rarely used on its own.  Full type polymorphism is more
  common, even if the type is decomposed and the parts used independently,
  as in this more general type for the \<choose> function:

  $\<choose> : \forallt{\kappa\ \beta} \<Set>\angles{\kappa\ \beta} \rightarrow \<NonDet>\ \beta$

\item
Qualifier polymorphism is common.  For example, here are types for
the \<length> method on strings and the addition operation:

@@

@@

$\<length>\ : \<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$

$\<length>\ :\<Det>\ \<String> \rightarrow \<Det> \<Int>$

The body of \<length> must typecheck at every possible instantiation of its
polymorphic function type.  At a use site of \<length>,
the most specific applicable instantiation is chosen.
Here is an example:

Suppose we have two variables $\<ni> : \<NonDet Int>$ and $\<di> : \<Det Int>$.
This table shows, for several method invocations, the 
most specific applicable instantiation for that invocation and the type of
the call expression:

\begin{tabular}{lll}
  Expression & instantiation & type \\
\<plus(ni, ni)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$ & \<NonDet> \\
\<plus(di, di)> & $\<Det>\ \<String> \rightarrow \<Det> \<Int>$ & \<Det> \\
\<plus(ni, di)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$ & \<NonDet> \\
\<plus(di, ni)> & $\<NonDet>\ \<String> \rightarrow \<NonDet> \<Int>$ & \<NonDet> \\
\end{tabular}

\smallskip

We adopt the convention that  polymorphism is not instantiated in ways
that would create invalid types.  For example, the \<length> and \<plus> polymorphic
functions would not be instantiated at $\kappa = \<OrderNonDet>.$
This makes no difference for these particular functions:  such an
instantiation would never be the most specific applicable one.
This is important for some of the type rules given in this paper, which
would otherwise need complex preconditions that recapitulate the type
well-formedness rules.

@@

\noindent
because such an instantiation would include the invalid return type \<OrderNonDet Int>.
This means that the only two instantiations are at 
$\kappa = \<NonDet>$ and $\kappa = \<Det>$.  An invocation on a
\<OrderNonDet> collection would choose the \<NonDet> instantiation and the
return type at that call site would be \<NonDet Int> rather than the
desired \<Det Int>.

\OurTypeSystem resolves this problem by introducing two operators over
polymorhic (type) variables, $\uparrow$ and $\downarrow$.
The $\uparrow$ operator converts \<OrderNonDet> to \<NonDet> and leaves the
other qualifiers unchanged.  The upward-pointing arrow is mnemonic for replacing
\<OrderNonDet> by something higher in the type hierarchy.
The $\downarrow$ operator is analogous, but it converts \<OrderNonDet> to
\<Det>, which is lower in the type hierarchy.
\Cref{fig:poly-resolutions} formalizes their behavior.

@@

\paragraph{Discussion of the type of \<length>}

In the type of \<first>, the $\uparrow$ is not strictly necessary.

Suppose that $\<myStrings> : \ptype{OrderNonDet List}{Det String}$.
The type of \<first(myStrings)> should be \<NonDet String>.

If the type of \<first> were

$\<first> : \forallt{\kappa} \kappa\ \ListKB \rightarrow \kappa\ \beta_e$

\noindent
then the instantiation

$\<first> : \<OrderNonDet>\ \ptype{List}{$\kappa$ String} \rightarrow \<OrderNonDet>\ \<String>$

\noindent
would be invalid.
At the call \<first(myStrings)>, the most specific instantiation would be

$\<first> : \<NonDet>\ \ptype{List}{$\kappa$ String} \rightarrow \<NonDet>\ \<String>$

\noindent making the type of the expression \<NonDet String> as desired.
We prefer the type

\noindent
because it is more explicit about the method's effect and more
instantiations, rather than requiring
a reader to reason about valid types.

The rules prevent some invocations.  Suppose we have the following variable:

$\<listOfLists> : \ptype{OND List}{\ptype{Det List}{Det String}}$

\noindent
Then the invocation \<first(listOfLists)> is illegal.
Its type would be \ptype{NonDet List}{Det String}, which is an invalid
type.

This restriction prevents programmers from writing some code, but we never
found this to be a problem in our case studies.
